Natural selection in the Health and Retirement Study

### 2023-09-22 hrs-selection.qmd
Rearrange a para. Add R file extrapolating to effect sizes of true pgs

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                                     \(x) svyglm_quiet(x, r_eth))

@@

@fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Bootstraps are used so as to make inferences from the sample of respondents (the different PGS are not a "sample" of anything).

@@

I next test part 2 of the theory more directly by interacting PGS with measures of education, income, marital status, and age at first birth. Education is years of education, split at 12 years. Income is respondent's mean wage income over all surveys, residualized on a full set of birth year dummies, and median-split. From here on I only use the white subsample: there are too few black respondents to be informative.

@@

@@

Why does the US data show fewer differences by socio-economic status (SES) than the UK? One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK. The regression coefficient of phenotypic educational attainment on fertility is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I. `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I. `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf. @goldscheider1969minority, @johnson1979minority, @yang2003big). And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated (see @fig-rlrs-edu). But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and by their different predictive power, so this hypothesis can only be speculative.

@@

tidy_pgs$white |>
  mutate(
    term = nice_names[term],
    term = fct_reorder(term, estimate)
  ) |>
  ggplot(aes(estimate, term)) +
    geom_vline(xintercept = 0) +
    geom_pointrange(aes(xmin = conf.low, xmax = conf.high), size = 0.3) +
    theme_light() + 
    labs(
      x = "", y = "",
      title = "Coefficients of polygenic scores on RLRS"
    ) +
    theme(
      axis.text.y = element_text(size = 7),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "none"
    )
```

>>>>

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                                     \(x) svyglm_quiet(x, r_eth))
  names(mod_pgs_edyrs[[eth]]) <- pgs

@@

@fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents (the different PGS are not a "sample" of anything).

@@

I next test part 2 of the theory by interacting PGS with measures of education, income, marital status, and age at first birth. Education is years of education, split at 12 years. Income is respondent's mean wage income over all surveys, residualized on a full set of birth year dummies, and median-split. From here on I only use the white subsample: there are too few black respondents to be informative.

@@

```{r}
#| label: stat-pgs-over-time
#| eval: false

# Plot changes in a couple of variables by decade
# Below: "Changes in PGS over time are small for both groups..."

rand$variables$birth_decade <- chop_width(rand$variables$rabyear, 10, 1920)
decades_adhd <- svyby(~ADHD_PGC17, ~ birth_decade + ethnicity, rand, 
                      FUN = svymean) 
decades_ea3 <- svyby(~EA3, ~ birth_decade + ethnicity, rand, 
                     FUN = svymean)

# example plot
ggplot(decades_ea3, aes(birth_decade, EA3, color = ethnicity, 
                        group = ethnicity)) + 
  geom_line() +
  geom_pointrange(aes(ymax = EA3 + 1.96 * se, ymin = EA3 - 1.96 * se))

Why does the US data show fewer differences by socio-economic status (SES) than the UK? One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK. The regression coefficient of phenotypic educational attainment on fertility is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I. `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I. `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf. @goldscheider1969minority, @johnson1979minority, @yang2003big). And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated (see @fig-rlrs-edu). But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and differences in the scores' predictive power, so this hypothesis can only be speculative. Looking within the sample, changes in PGS by birth year are small for both groups, and are probably mostly driven by selective mortality.

@@

@@

tidy_pgs$white |>
  mutate(
    term = nice_names[term],
    term = fct_reorder(term, estimate)
  ) |>
  ggplot(aes(estimate, term)) +
    geom_vline(xintercept = 0) +
    geom_pointrange(aes(xmin = conf.low, xmax = conf.high), size = 0.3,
                    color = "steelblue3") +
    theme_light() + 
    labs(
      x = "", y = "",
      title = "Coefficients of polygenic scores on RLRS"
    ) +
    theme(
      axis.text.y = element_text(size = 7),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_blank(),
      legend.position = "none"
    )
```
